Parametricity Consequences of GADTs Representations

### 2021-02-15 20-gadts/GADT-story.tex
work on gadt-story Sec 2.2

Primitive representations capture in syntax the fact that ADTs can be
considered as fixpoints. For example $\mathsf{List\,a}$ does indeed
satisfy the fixpoint equation indicated above, i.e., is indeed such
that
\[\mathsf{List\,a} = \mathsf{1 + (a \times List\,a)}\] 
since every element of $\mathsf{List\,a}$ is either empty or is
obtained by $\mathsf{Cons}$ing an element of type $\mathsf{a}$ onto an
already-existing structure of type $\mathsf{List\,a}$. (The fixpoint
equation is just a rewriting of the Haskell data type declaration for
$\mathsf{List\,a}$.) The ADT $\mathsf{Tree\,a}$ can similarly be seen
to be a solution to the fixpoint equation
\[\mathsf{Tree\,a} = \mathsf{a + (Tree\,a \times a \times Tree\,a)}\] 
determined by its Haskell data type declaration.

@@

GADTs are useful in such situations precisely because generalizing the
type arguments in the instances of a GADT appearing as the return
types of its data constructors entails a strong connection between a
GADTs shape and the data it contains.  With ADTs, we first choose the
shape of the container and then fill that container with data of
whatever type we like. Critically, the choice of shape is independent
of the data to be stored. With GADTs, however, the shape of the
``container'' may actually {\em depend} on (the type of) the data to
be contained. On other words, the shape and the data are not
independent! As a result, it may not always be possible to modify a
GADT's data while retaining its shape because changing (the type of)
the data may actually change the shape.

To determine the possible shapes of a GADT's container we have to
pattern-match on the type of the data to be contained. For this, our
GADT calculus must support an equality type $\mathsf{Eq}$ that is
inhabited when its two type arguments are the same and is uninhabited
otherwise, i.e., is the syntactic reflection of semantic equality
$\Eq$. Such a type can either be taken as primitive, as in
\[\begin{array}{l}
\mathsf{data\, Eq :: * \to * \to *\,where}\\
\mathsf{\;\;\;\;\;\;\;\;Refl ::\, Eq\,a\,a}\\
\end{array}\]
\noindent
or
%, in a higher-kinded calculus such as $F\omega$,
represented by its own Church encoding
\[\mathsf{Eq\,t_1\,t_2} = \mathsf{\forall f.\, (\forall a.\,
  f\,a\,a) \to f\,t_1\,t_2}\]
Using $\mathsf{Eq}$, the GADT $\mathsf{Seq}$ can be written as
%\[\mathsf{Seq\,a} = \mathsf{a + Eq\,a\,(b \times c) \times Seq\,b
%  \times Seq\,c} \] 
\[\begin{array}{l}
\mathsf{data\, Seq :: * \to *\,where}\\
\mathsf{\;\;\;\;\;\;\;\;Const ::\, a \to Seq\,a}\\
\mathsf{\;\;\;\;\;\;\;\;Pair\,\,\,\,\, ::\, Eq\,a\,(b \times c) \to
  Seq \,b \to Seq\,c \to Seq\,a}\\ 
\end{array}\]

This rewriting of $\mathsf{Seq}$'s definition shows that it can be
seen as a fixpoint $\mu H$ of the higher-order function $H$ defined on
objects by $H\,F\,a = a + Eq\,a\,(b \times c) \times F\,b \times
F\,c$. We note, however, that $Eq\,a$ cannot be made a functor, or,
equivalently, that $\mathsf{Eq\,a}$ cannot be an instance of Haskell's
$\mathsf{Functor}$ class. Indeed, if $\mathsf{Eq\,a}$ supported a
function $\mathsf{map_{Eq\,a} :: (b \to c) \to Eq\,a\,b \to Eq\,a\,c}$
then defining
\[\begin{array}{l}
\mathsf{eqElim :: Eq\, a\, b \to b \to a}\,{\color{red} wasEq\,a\,b\to b\to a}\\
\mathsf{eqElim\, Refl\, x = x} 
\end{array}\]
would allow us to construct the following element of the type
$\mathsf{Void}$:
\[\begin{array}{l}
\mathsf{emptyFromMapEqa :: (\forall b\, c.\, (b \to c) \to Eq\, a\,
  b \to Eq\, a \,c) \to Void}\\ 
\mathsf{emptyFromMapEqa \,map_{Eq\,a} = eqElim\, (map_{\Eq\,a}\,
  absurd\, Refl)\, ()}  
\end{array}\]
Of course this is impossible.  It follows immediately that
$\mathsf{Seq}$ cannot support a map function either. This is already
evident from the Haskell definition of $\mathsf{Seq}$, though. The
clause of $\mathsf{map}$ for the $\mathsf{Pair}$ constructor should
feed $\mathsf{map}$ a function $\mathsf{f :: (b \times c) \to e}$ and
a term of the form $\mathsf{Pair \,t_1\,t_2}$ for $\mathsf{x :: b}$
and $\mathsf{y :: c}$, and produce a term $\mathsf{Pair\,t_1'\,t_2'}$
for some appropriately typed terms $\mathsf{t_1'}$ and
$\mathsf{t_2'}$. But it is not clear how to achieve this since
$\mathsf{e}$ need not necessarily be a product type. And even if
$\mathsf{e}$ {\em were} known to be of the form $\mathsf{w \times z}$,
then we still wouldn't necessarily have a way to produce data of type
$\mathsf{w \times z}$ from only $\mathsf{f :: b \times c \to w \times
  z}$ and $\mathsf{x}$ and $\mathsf{y}$ unless we knew, e.g., that
$\mathsf{f}$ was a pair of functions.

@@

The primitive representations of {\color{blue} polynomial} GADTs
capture in syntax the fact that GADTs can be considered as fixpoints
of higher-order functors. The use of {\em higher-order functions} is
essential, since the type arguments to the GADT being defined are not
uniform across all of its instances in the types of its data
constructors; the use of higher-order funct{\em ors} is essential, as
for ADTs, to the very existence of the fixpoints being
computed~\cite{tfca}.

Consider, for example, the GADT $\mathsf{Seq}$ from above.  Because
its type argument varies in the instances of $\mathsf{Seq}$ appearing
in the types of its data constructor $\mathsf{Pair}$, $\mathsf{Seq}$
cannot be characterized by any first-order fixpoint equation. It can,
however, be viewed as a solution to the higher-order fixpoint equation
{\color{blue} syntax or semantics?}
\[\mathsf{H\,f\,a = a \,+\,Lan_{\lambda a\,b.\,a \times b}\,
  \lambda a\,b.\,f\,a \times f\,b}\] where $\mathsf{Lan_k\, g}$ is the
syntactic reflection of the {\em left Kan extension} $\Lan_K \,G$ of
the functor $G$ interpreting $\mathsf{g}$ along the functor $K$
interpreting $\mathsf{k}$~\cite{mac71}.

Other GADTs are similarly characterizable by higher-order functors.
The difference between ADTs and GADTs can thus be seen as a difference
in {\em orders} of their underlying functors. {\color{blue} GADTs
  make essential use of higher-order functors.}

@@

Certain ``polynomial'' GADTs correspond to polynomial ADTs. So we
expect the same behaviors.

Polynomial nested types are actually the GADTs that mimic polynomial
ADT behavior.

@@

@@

Functorial comnpletion doesn't give nested types

@@

>>>>

Primitive representations capture in syntax the fact that ADTs can be
considered as fixpoints. For example $\mathsf{List\,a}$ does indeed
satisfy the fixpoint equation indicated above, i.e., is indeed such
that
\[\mathsf{List\,a} = \mathsf{1 + (a \times List\,a)}\] 
since every element of $\mathsf{List\,a}$ is either empty or is
obtained by $\mathsf{Cons}$ing an element of type $\mathsf{a}$ onto an
already-existing structure of type $\mathsf{List\,a}$. (The fixpoint
equation is just a rewriting of the Haskell data type declaration for
$\mathsf{List\,a}$.) The ADT $\mathsf{Tree\,a}$ can similarly be seen
to be a solution to the fixpoint equation
\[\mathsf{Tree\,a} = \mathsf{a + Tree\,a \times a \times Tree\,a}\] 
determined by its Haskell data type declaration.

@@

GADTs are useful in such situations precisely because generalizing the
type arguments in the instances of a GADT appearing as the return
types of its data constructors entails a strong connection between a
GADTs shape and the data it contains.  With ADTs, we first choose the
shape of the container and then fill that container with data of
whatever type we like. Critically, the choice of shape is independent
of the data to be stored. With GADTs, however, the shape of the
container may actually {\em depend} on (the type of) the data to be
contained. On other words, the shape and the data are not independent!
For example, while $\mathsf{Const}$ can create data of any shape
$\mathsf{Seq\,a}$, $\mathsf{Pair}$ can produce data of shape
$\mathsf{Seq\,a}$ where $\mathsf{a}$ is a pair type. As a result, it
may not always be possible to modify a GADT's data while retaining its
shape because changing (the type of) the data may actually change the
shape.

To determine the possible shapes of a GADT's container we have to
pattern-match on the type of the data to be contained. For this, our
GADT calculus must support an equality type $\mathsf{Eq}$ that is
inhabited when its two type arguments are the same and is uninhabited
otherwise, i.e., is the syntactic reflection of semantic equality
function $Eq$. Such a type can either be taken as primitive, as in
\[\begin{array}{l}
\mathsf{data\, Eq :: * \to * \to *\,where}\\
\mathsf{\;\;\;\;\;\;\;\;Refl ::\, Eq\,a\,a}\\
\end{array}\]
\noindent
or
represented by its own Church encoding
\[\mathsf{Eq\,t_1\,t_2} = \mathsf{\forall f.\, (\forall a.\,
  f\,a\,a) \to f\,t_1\,t_2}\] Using $\mathsf{Eq}$, the GADT
$\mathsf{Seq}$ can then be represented as its Church encoding
         {\color{red}
\[\mathsf{Seq\,a} = \mathsf{\forall f.\, (\forall a.\,a \to f\, a)
  \to (\forall a\,b\,c.\,Eq\,a\,(b\times c) \to f\,b \to f\,c \to
  f\,a) \to f\,a}\]}
in a higher-kinded calculus such as $F_\omega$.

Importantly, $Eq\,a$ cannot be made into a functor, or, equivalently,
$\mathsf{Eq\,a}$ cannot be made an instance of Haskell's
$\mathsf{Functor}$ class: if $\mathsf{Eq\,a}$ supported a function
$\mathsf{map_{Eq\,a} :: (b \to c) \to Eq\,a\,b \to Eq\,a\,c}$ then
defining
\[\begin{array}{l}
\mathsf{eqElim :: Eq\, a\, b \to b \to a}\,{\color{red} was Eq\,a\,b\to a\to b}\\
\mathsf{eqElim\, Refl\, x = x} 
\end{array}\]
would allow us to construct the element 
\[\begin{array}{l}
\mathsf{emptyFromMapEqa :: (\forall b\, c.\, (b \to c) \to Eq\, a\,
  b \to Eq\, a \,c) \to Void}\\ 
\mathsf{emptyFromMapEqa \,map_{Eq\,a} = eqElim\, (map_{\Eq\,a}\,
  absurd\, Refl)\, ()}  
\end{array}\]
of type $\mathsf{Void}$, which is impossible. It follows immediately
from this contradiction that $\mathsf{Seq}$ cannot support a map
function either. This is already intuitively evident from the Haskell
definition of $\mathsf{Seq}$, though. Indeed, the clause of
$\mathsf{map}$ for the $\mathsf{Pair}$ constructor should feed
$\mathsf{map}$ a function $\mathsf{f :: (b \times c) \to e}$ and a
term of the form $\mathsf{Pair \,t_1\,t_2}$ for $\mathsf{x :: b}$ and
$\mathsf{y :: c}$, and produce a term $\mathsf{Pair\,t_1'\,t_2'}$ for
some appropriately typed terms $\mathsf{t_1'}$ and
$\mathsf{t_2'}$. But it is not clear how to achieve this since
$\mathsf{e}$ need not necessarily be a product type. And even if
$\mathsf{e}$ {\em were} known to be of the form $\mathsf{w \times z}$,
then we still wouldn't necessarily have a way to produce data of type
$\mathsf{w \times z}$ from only $\mathsf{f :: b \times c \to w \times
  z}$ and $\mathsf{x}$ and $\mathsf{y}$ unless we knew, e.g., that
$\mathsf{f}$ was a pair of functions.

@@

The primitive representations of {\color{blue} polynomial} GADTs
capture in syntax the fact that GADTs can be considered as fixpoints
of higher-order functors. Syntactically, the use of higher-order
funct{\em ions} is essential, since the type arguments to the GADT
being defined are not uniform across all of its instances in the types
of its data constructors. Semantically, the use of higher-order
funct{\em ors} is essential, as for ADTs, to the very existence of the
fixpoints being computed~\cite{tfca}.
%We will consider GADTs'
%semantics in more detail in Section~\ref{sec:GADT-semantics}.

Consider, again, the GADT $\mathsf{Seq}$ from the start of
Section~\ref{sec:gadts}. Because its type argument varies in the
instances of $\mathsf{Seq}$ appearing in the types of its data
constructor $\mathsf{Pair}$, $\mathsf{Seq}$ cannot be characterized by
any first-order fixpoint equation. It can, however, be viewed as a
solution to the higher-order fixpoint equation
\[H\,f\,a = a \,+\,Lan_{\lambda a b.\,a \times b}\, \lambda
a b.\,f\,a \times f\,b\] 
where $\Lan_K \,G$ is the 
%$Lan_k\, g$ is the syntactic reflection of the
{\em left Kan extension}~\cite{mac71}
%$\Lan_K \,G$
of the functor $G$
%interpreting $\mathsf{g}$
along the functor $K$.
%interpreting $\mathsf{k}$~\cite{mac71}.
The left Kan extension $\Lan_K \,G$ is the smallest functor
approximating/extending $K$ and coinciding with $G$ {\color{blue}
  better explanation}.

LKE gives the functorial completion of the GADT.

%written as
%%\[\mathsf{Seq\,a} = \mathsf{a + Eq\,a\,(b \times c) \times Seq\,b
%%  \times Seq\,c} \] 
%\[\begin{array}{l}
%\mathsf{data\, Seq :: * \to *\,where}\\
%\mathsf{\;\;\;\;\;\;\;\;Const ::\, a \to Seq\,a}\\
%\mathsf{\;\;\;\;\;\;\;\;Pair\,\,\,\,\, ::\, Eq\,a\,(b \times c) \to
%  Seq \,b \to Seq\,c \to Seq\,a}\\ 
%\end{array}\]
%
%The above rewriting of $\mathsf{Seq}$'s definition shows that it can
%be seen as a fixpoint $\mu H$ of the higher-order function $H$ defined
%on objects by $H\,F\,a = a + Eq\,a\,(b \times c) \times F\,b \times
%F\,c$.

{\color{blue} Need syntax that reflects semantic Lan.}

%Other GADTs are similarly characterizable by higher-order functors.
%{\color{blue} GADTs make essential use of higher-order functors.}

@@

Certain polynomial GADTs correspond to polynomial ADTs. So we expect
the same behaviors.  Polynomial nested types are actually the GADTs
that mimic polynomial ADT behavior.

@@

{\color{blue} OMIT THIS SECTION?}

@@

Functorial completion doesn't give nested types

@@

Bob requires map functions for fixpoint but then doesn't use them.
